Why do we Eval?\\[2mm]\Large A large-scale study of Eval usage in R

### 2020-11-27 inst/paper/main.tex
sec 2

\title{Why do we Eval?\\[2mm]\Large A large-scale study of Eval usage in R}
\titlerunning{Eval in R}

@@

\vspace{2mm}\noindent\emph{Trouble in Paradise.} Reflective facilities
hinder most attempt to reason about the code using them, as well as to
apply meaning-preserving program transformations. In practice this
means that static analysis techniques will loose so much precision as
to become pointless and that anything but the most trivial, local,
compiler optimizations will be unsound. Furthermore the addition of
new code --code that could have been obtained from a network
connection-- as the program is running is a security vulnerability
waiting to happen.

@@

\vspace{2mm}\noindent\emph{The What and How.}  One significant benefit
of choosing R is that every package in the CRAN repository is curated
and comes with examples of typical usage. This gives us a large code
base that we can analyze dynamically.  To observe \eval we built a
two-level monitoring infrastructure: we can monitor R programs by
instrumentation --this gives us access to many user-visible properties
of R programs-- but we can also monitor the inner working of the R
interpreter --this allows us to capture details of execution that are
not exposed at the source level. Dynamic analysis is of course limited
in that it can only observe behaviors that are triggered by the
particular inputs passed to the program. Luckily, CRAN libraries come
with multiple input datasets and use-cases.

@@

\noindent
These design choices impact our ability to reason about code. More
expressive power for \eval translates directly into additional
restriction for program analysis and transformation tools. The reason
Julia has restricts \eval to execute in top level environment is that
they wanted to preserve their compiler's ability to optimize
functions~\cite{oopsla18a}. What they wanted to prevent was the case
where \eval changed the type of local variables as this would require
recompilation of the method (or prevent optimization). Julia even adds
a versoning mechanism, called world age, to ensure that code added
during an \eval does not invalidate optimizations such as
inlining~\cite{oopsla20a}.

R's ability to affect any one of the callers of a function is even
more invasive as the impact of \eval is non-local. Such unfettered
power is difficult to deal with.

There are two outstanding differences that set the design of \eval in
R apart from JavaScript. On the usability side, \eval does not take a
string, but an AST and an environment. This is slightly more involved
than just taking a string and may discourage some casual users. The
second substantial difference is the control that R gives over the
choice of environment in which the AST is evaluated in. Beside
traditional choices such as the top level or current lexical
environment, R allows programmers to choose a fresh, empty,
environment or any environment on the call stack. This is a
significant increase in expressive power as functions that are
otherwise not aware of the use of \eval may be deeply impacted.

R supports reflection and meta-programming. The function
\c{substitute(e,envir)} yields the parse tree of the expression \c{e}
after performing substitutions defined by the bindings in the
\c{envir} parameter.

@@

\noindent
R allows programmatic manipulation of parse trees, which are themselves
first class objects. They are evaluated using the \c{eval()}
function. There are three alternative forms  of \eval in the \c{base} package:
\c{evalq}, \c{eval.parent} and \c{local}.

\eval function in R takes three arguments as shown below.
\begin{lstlisting}
 eval(expr,
      envir = parent.frame(),
      enclos = if(is.list(envir) || is.pairlist(envir))
               parent.frame() else baseenv())
\end{lstlisting}

\c{expr} is the value to be evaluated, \c{envir}
is the environment in which it is to be evaluated, and \c{enclos} is
relevant when \c{envir} is a list or a data frame, then it specifies where R
looks for objects not found in \c{envir}. This can be \c{NULL} (interpreted
as the base package environment, \c{baseenv()}) or an environment.

When called, \eval evaluates \c{expr} in the environment specified by
\c{envir} and returns the computed value. If \c{envir} is missing, then the
default is \c{parent.frame()} (the environment where the call to \eval was
made).  Values to be evaluated can be of types \c{call} (AST for function
calls) or \c{expression} (AST nodes for arbitrary expressions), \c{name}
(when the symbol is looked up in the current scope and its binding is
evaluated), promise, or any of the basic types such as vectors and
environments (which are returned unchanged).

The \c{evalq} form is equivalent to \c{eval(quote(expr), ...)}.  Quoting
that argument prevents if from being evaluated in the current environment.
In the following code snippet, variable \c{r} is first looked up in
the current scope with \eval whereas it is looked up in \c{env} with
\c{evalq}.

\begin{lstlisting}
> r <- 5
> env <- new.env()
> env$r <- 9
> eval(r, envir = env)
[1] 5
> evalq(r, envir = env)
[1] 9
\end{lstlisting}
% $

\c{eval.parent(expr, n)} is a shorthand for \c{eval(expr, parent.frame(n))}.
\c{local} evaluates an expression in a local environment. It is equivalent
to evalq except that its default argument creates a new, empty
environment. This is useful to create anonymous recursive functions and as a
kind of limited namespace feature since variables defined in the environment
are not visible from the outside.

>>>>

\title{Why do we Eval?\\[2mm]\Large A large-scale study of Eval usage in R}
\titlerunning{Why Eval (in R)?}

@@

\vspace{2mm}\noindent\emph{Trouble in Paradise.} Reflective facilities
hinder most attempt to reason about the code using them, as well as to
apply meaning-preserving program transformations. In practice this
means that static analysis techniques will loose so much precision as
to become pointless and that anything but the most trivial, local,
compiler optimizations will be unsound. Furthermore the addition of
new code --- code that could have been obtained from a network
connection --- as the program is running is a security vulnerability
waiting to happen.

@@

\vspace{2mm}\noindent\emph{The What and How.}  One significant benefit
of choosing R is that every package in the CRAN repository is curated
and comes with examples of typical usage. This gives us a large code
base that we can analyze dynamically.  To observe \eval we built a
two-level monitoring infrastructure: we can monitor R programs by
instrumentation --- this gives us access to many user-visible properties
of R programs --- but we can also monitor the inner working of the R
interpreter --- this allows us to capture details of execution that are
not exposed at the source level. Dynamic analysis is of course limited
in that it can only observe behaviors that are triggered by the
particular inputs passed to the program. Luckily, CRAN libraries come
with multiple input datasets and use-cases.

@@

\noindent
These design choices impact our ability to reason about code. More
expressive power for \eval translates directly into additional
restriction for program analysis and transformation tools. The reason
Julia has restricts \eval to execute in top level environment is that
they wanted to preserve their compiler's ability to optimize
functions~\cite{oopsla18a}. What they wanted to prevent was the case
where \eval changed the type of local variables as this would require
recompilation of the method (or prevent optimization). Julia even adds
a versoning mechanism, called world age, to ensure that code added
during an \eval does not invalidate optimizations such as
inlining~\cite{oopsla20a}. More permissive languages such as R are
much harder to deal with from a compiler's perspective.

\medskip\noindent{\it Eval in R.} There are four reflective functions
that take abstract syntax trees (called expressions), parse it as
code, and evaluate it. These functions are \c{eval}, \c{evalq},
\c{eval.parent}, and \c{local}. Unlike other languages, R's \eval does
not take a string, but rather an expression. As this is slightly more
involved, it may discourage some casual users. There are two ways to
obtain an expression. Given a string, the \c{parse()} function retuns
an unevaluated expression. The other way is to use the function
\c{substitute(e,envir)}, it yields the parse tree of the expression
\c{e} after performing substitutions defined by the bindings in the
\c{envir} parameter.

@@

\noindent
This function is widely used in conjunction with \eval, as it can
leverage R's laziness for metaprogramming. R allows programmatic
manipulation of parse trees, which are themselves first class objects.
Expressions are evaluated using \eval (or one its variants). The \eval
function takes three arguments:

\begin{lstlisting}
 eval(expr,
      envir = parent.frame(),
      enclos = if(is.list(envir) || is.pairlist(envir))
               parent.frame() else baseenv())
\end{lstlisting}

\noindent
This code is part of the definition of \eval, it specifies that the
function takes an \c{expr} which is the value to be evaluated,
\c{envir} is the environment in which evaluation happes, and
\c{enclos} is relevant when \c{envir} is a list or a data frame, then
it specifies where R looks for objects not found in \c{envir}. This
can be \c{NULL} (interpreted as the base package environment,
\c{baseenv()}) or an environment. Thus, the expression \c{expr}
evaluates in the environment \c{envir} and returns the computed value.
If \c{envir} is missing, then the default is \c{parent.frame()} (the
environment where the call to \eval was made). Values to be evaluated
can be of types \c{call} (function calls) or \c{expression} (arbitrary
expressions), \c{name} (when the symbol is looked up in the current
scope and its binding is evaluated), promise, or any of the basic
types such as vectors and environments (which are returned unchanged).

The \c{evalq} form is equivalent to \c{eval(quote(expr), ...)}.
Quoting that argument prevents if from being evaluated in the current
environment. \c{eval.parent(expr, n)} is a shorthand for \c{eval(expr,
  parent.frame(n))}. \c{local} evaluates an expression in a local
environment. It is equivalent to evalq except that its default
argument creates a new, empty environment. This is useful to create
anonymous recursive functions and as a kind of limited namespace
feature since variables defined in the environment are hidden.

The second substantial difference is the control that R gives over the
choice of environment in which the AST is evaluated in. Beside
traditional choices such as the top level or current lexical
environment, R allows programmers to choose a fresh, empty,
environment or any environment on the call stack. This is a
significant increase in expressive power as functions that are
otherwise not aware of the use of \eval may be deeply impacted.
